---
title: "Submission_2_103"
output: pdf_document
---

## R Markdown

```{r}
#Convert genes into gene series
library(readr)
setwd("/Users/deepthi/Documents/GitHub/Final_103_Project_LDG")
# read in genes file
Genes<- read_csv("QBS103_GSE157103_genes.csv")
#read in Gene Series file
Genes_Series <- read_csv("QBS103_GSE157103_series_matrix.csv")

#transpose gene data to put participant id as rows and genes as columns 
TGenes <-as.data.frame(t(Genes))
# rearrange table to make gene names as column names instead of row names
names(TGenes) <- TGenes[1,]
#remove repetitive first column with genes
TGenes <- TGenes[-1,]
#create a column with the row names( participant ids)to merge with gene series file
TGenes$participant_id <- row.names(TGenes)
# Merge Gene and Gene Series tables by participant_id 
MergedGenes <- merge(TGenes,Genes_Series, by = 'participant_id', all = TRUE) 
row.names(MergedGenes) <- MergedGenes$participant_id

# change ICU status values for proper formatting
MergedGenes$icu_status[MergedGenes$icu_status =='no'] = 'No'
MergedGenes$icu_status[MergedGenes$icu_status =='yes'] = 'Yes'
# change sex values for proper formatting
MergedGenes$sex[MergedGenes$sex=='female'] = 'Female'
MergedGenes$sex[MergedGenes$sex =='male'] = 'Male'
MergedGenes$sex[MergedGenes$sex =='unknown'] = 'Unknown'
```
Generate a table formatted in LaTeX of summary statistics for all the covariates you
looked at and 2 additional continuous (3 total) and 1 additional categorical variable (3
total). (5 pts)
o Stratifying by one of your categorical variables
o Tables should report n (%) for categorical variables
o Tables should report mean (sd) or median [IQR] for continuous variables

Generate final histogram, scatter plot, and boxplot from submission 1 (i.e. only for your
first gene of interest) incorporating all feedback from your presentations (5 pts)
Generate a heatmap (5 pts)
o Heatmap should include at least 10 genes
o Include tracking bars for the 2 categorical covariates in your boxplot
o Heatmaps should include clustered rows and columns

Going through the documentation for
ggplot2, generate a plot type that we did not
previously discuss in class that describes your data in a new and unique way

```{r}

# load in libraries
library(dplyr)
library(tidyverse)

# create a subiset of df with necesarry continours and categorical variables 
Genes_CoVariate_Data_3_3 <- MergedGenes %>%
   select(age, sex,icu_status,disease_status,`hospital-free_days_post_45_day_followup`,`ventilator-free_days`) 
# convert the >89 age to 90 years old
Genes_CoVariate_Data_3_3$age[Genes_CoVariate_Data_3_3$age== ">89"] <- 90

#convert continuous column to numeric type to numeric data
Genes_CoVariate_Data_3_3$age <- as.numeric(Genes_CoVariate_Data_3_3$age)
Genes_CoVariate_Data_3_3$`hospital-free_days_post_45_day_followup` <- as.numeric( Genes_CoVariate_Data_3_3$`hospital-free_days_post_45_day_followup`)
Genes_CoVariate_Data_3_3$`ventilator-free_days` <- as.numeric( Genes_CoVariate_Data_3_3$`ventilator-free_days`)


#check distribution of each of the continuous variables to decide whether to use mean/sd or median/IQR
Age <- ggplot(data = Genes_CoVariate_Data_3_3,aes(x = age))+
  geom_histogram(binwidth = 5)+  
  labs(x = 'Age',y = 'Frequency') +
  theme_classic()
print(Age)

hospital_free_days <- ggplot(data = Genes_CoVariate_Data_3_3,aes(x = `hospital-free_days_post_45_day_followup`))+
  geom_histogram(binwidth = 5) + 
  labs(x ='hospital-free_days_post_45_day_followup', y = 'Frequency') +
  theme_classic()
print(hospital_free_days)

ventilator_free_days <- ggplot(data = Genes_CoVariate_Data_3_3, aes(x = `ventilator-free_days`))+
    geom_histogram(binwidth = 5)+ 
    labs(x ='ventilator-free_days',y = 'Frequency')+
    theme_classic()
print(ventilator_free_days)

```

Generate a table formatted in LaTeX of summary statistics for all the covariates you
looked at and 2 additional continuous (3 total) and 1 additional categorical variable (3
total). (5 pts)
o Stratifying by one of your categorical variables
o Tables should report n (%) for categorical variables
o Tables should report mean (sd) or median [IQR] for continuous variables

```{r}


library(kableExtra)
library(table1)

#Format column names and factor values to create neat tables 

# change col names
names(Genes_CoVariate_Data_3_3) <- c('Age',"Sex","ICU_Status", "Disease_Status","Hospital_Free_Days_Post_45_Day_Followup", "Ventilator_Free_Days")

# change disease status values 
Genes_CoVariate_Data_3_3$disease_status[Genes_CoVariate_Data_3_3$disease_status =='disease state: non-COVID-19'] = 'Non-COVID-19'
Genes_CoVariate_Data_3_3$disease_status[Genes_CoVariate_Data_3_3$disease_status =="disease state: COVID-19"] = 'COVID-19'

# change ICU status values 
Genes_CoVariate_Data_3_3$ICU_Status[Genes_CoVariate_Data_3_3$ICU_Status =='no'] = 'No'
Genes_CoVariate_Data_3_3$ICU_Status[Genes_CoVariate_Data_3_3$ICU_Status =='yes'] = 'Yes'

# change sex values 
Genes_CoVariate_Data_3_3$Sex[Genes_CoVariate_Data_3_3$Sex =='female'] = 'Female'
Genes_CoVariate_Data_3_3$Sex[Genes_CoVariate_Data_3_3$Sex =='male'] = 'Male'
Genes_CoVariate_Data_3_3$Sex[Genes_CoVariate_Data_3_3$Sex =='unkown'] = 'Unknown'


# create a function that specifies what parameters to include for continuous values in the table1 generated summary statistics table 
## documentation used to generate function : https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html
render_Calc<- function(x, namecol, ...) {
    if (!is.numeric(x)) return(render.categorical.default(x))
    what <- switch(namecol,
        # keep mean for age and median for hospital and ventilator free days based on distribution of values 
        Age = "Mean (SD)",
        Hospital_Free_Days_Post_45_Day_Followup = "Median [Min, Max]",
        Ventilator_Free_Days = "Median [Min, Max]")
    parse.abbrev.render.code(c("", what))(x)
}


# create a summary table use tableone package to get summary statistics and include render_Calc() function to specify format
Sum_Tab <-table1(~Age+ `Hospital_Free_Days_Post_45_Day_Followup`+ `Ventilator_Free_Days` + ICU_Status+ Disease_Status| Sex, data=Genes_CoVariate_Data_3_3, render.missing = NULL, overall= F, render= render_Calc)
Sum_Tab


# use kable() to generate latex format of summary statistics table 
tab <- kable(x = Sum_Tab, caption = 'Summary Table',
             format = 'latex',booktabs = T,
            # col.names = c("Female", "Male","Unknown", "Overall"),
             align = c('l','r'),escape = T) %>%
  add_indent(positions = c(3,4),level_of_indent = 1)
tab

```

Generate final histogram, scatter plot, and boxplot from submission 1 (i.e. only for your
first gene of interest) incorporating all feedback from your presentations (5 pts)

```{r}
library(dplyr)
library(ggplot2)

# function to generate final histogram, scatter plot, and boxplot from submission 1
GenePlots <- function(df, geneName, Cont, Cat1, Cat2) {
  # Pull out gene expression data, continuous covariate, and categorical data
  geneName <- geneName[[1]]  # Assumes geneName is a single-element list
  Genes_CoVariate_Data <- df %>%
    select(all_of(c(geneName, Cont, Cat1, Cat2))) %>%
    as.data.frame()
  # Convert gene expression column to numeric
  Genes_CoVariate_Data[[geneName]] <- as.numeric(Genes_CoVariate_Data[[geneName]])
  
  # Plot histogram of gene expression data
  histograms <- hist(Genes_CoVariate_Data[[geneName]], 
      main = paste(geneName, 'Gene Expression Data'), xlab = 'Gene Expression', col = "light blue")
  print(histograms)
  # Replace age values greater than 89 with 90
  Genes_CoVariate_Data[[Cont]][Genes_CoVariate_Data[[Cont]] == ">89"] <- 90
  # Convert continuous column to numeric
  Genes_CoVariate_Data[[Cont]] <- as.numeric(Genes_CoVariate_Data[[Cont]])
  
  # Create age groups
  Genes_CoVariate_Data$AgeGroup <- cut(Genes_CoVariate_Data[[Cont]],
                                        breaks = c(0, 30, 40, 50, 60, 70, 80, 90),
                                        labels = c('Under 30', '30-40', '40-50', '50-60', '60-70', '70-80', '>80'))
  
  # Plot scatter plot of gene expression by age
  Scatterplot <- ggplot(Genes_CoVariate_Data, aes(x = !!sym(Cont), y = !!sym(geneName), color = AgeGroup)) +
    geom_point() +
    labs(title = paste('Gene Expression of', geneName, 'by Age'),
         x = 'Age (yrs)',
         y = paste(geneName, 'Gene Expression'),
         color = "Age Group") +
    theme_classic()
  print(Scatterplot)
  
# add color pallete to color boxplots
  colorPalette = c('light blue',' maroon')

# create box plots to show Gene Expression of ABCA3 by Sex and ICU Status
  boxPlots <- ggplot(Genes_CoVariate_Data, aes(x = !!sym(Cat1), y=!!sym(geneName), fill= !!sym(Cat2)) )+ 
    # Add box plot
    geom_boxplot() +
    # add colors
    scale_fill_manual(values= colorPalette) +
    # Change axis labels
    labs(title =paste('Gene Expression of',geneName,'by Sex and ICU Status'), x = 'Sex',y = 'Gene Expression',fill = 'ICU Status') +
    theme_classic()
  print(boxPlots)}

# specify gene name 'ABCA3'
GeneNames <- c('ABCA3')
#run Gene Plots function
GenePlots(MergedGenes, GeneNames, 'age', 'sex', 'icu_status')

```

Generate a heatmap (5 pts) 
o Heatmap should include at least 10 genes
o Include tracking bars for the 2 categorical covariates in your boxplot
o Heatmaps should include clustered rows and columns


``` {r}
library(pheatmap)

# Prepare annotation data

# change ICU status values for proper formatting
MergedGenes$icu_status[MergedGenes$icu_status =='no'] = 'No'
MergedGenes$icu_status[MergedGenes$icu_status =='yes'] = 'Yes'
# change sex values for proper formatting
MergedGenes$sex[MergedGenes$sex=='female'] = 'Female'
MergedGenes$sex[MergedGenes$sex =='male'] = 'Male'
MergedGenes$sex[MergedGenes$sex =='unknown'] = 'Unknown'


# save annotation data extracting sex and ice_status col info 
annotation_rows<- data.frame(
  Sex = MergedGenes$sex,
  ICU_Status = MergedGenes$icu_status
)
# make row names of original table and annotation_rows table same 
rownames(annotation_rows) <- rownames(MergedGenes)

#create a color list for each of the values in in sex
color1 <- c( "Male" = "blue","Female" = "pink", "Unknown" = "gray")
#create a color list for each of the values in in icu)status
color2 <- c("Yes" = 'yellow', "No" = 'green')

# assign color list to categorical varaibles 
annotation_colorss <- list(
  Sex = color1,
  ICU_Status = color2
)

#Use apply function to convert all 10 selected gene columns to numeric format
Genes_numeric_df<- as.data.frame(apply(MergedGenes[,31:41],MARGIN = 2, function(x) {as.numeric(x)}))

# make rown ames of original table and 10 genes subset table same 
row.names(Genes_numeric_df) <- rownames(MergedGenes)

# generate heatmap and turn clustering and add annotations for sex and icu_status
pheatmap(mat=Genes_numeric_df,
         show_rownames = F,
         cluster_rows = T,
         cluster_cols = T,
         annotation_row = annotation_rows,
         annotation_colors = annotation_colorss)

```
Going through the documentation for
ggplot2, generate a plot type that we did not
previously discuss in class that describes your data in a new and unique way

```{r}

# create a 
Genes_CoVariate_Data <- MergedGenes %>%
  select(ABCA3,age,sex,icu_status) 

#install.packages("ggbeeswarm")
library(ggbeeswarm)
Genes_CoVariate_Data$ABCA3 <- as.numeric(Genes_CoVariate_Data$ABCA3 )
# Create a swarm plot with 
ggplot(Genes_CoVariate_Data, aes(x = interaction(sex, icu_status), y = ABCA3, color = interaction(sex, icu_status))) +
  geom_beeswarm(alpha = 1) +
  labs(title = "Beeswarm Plot of ABCA3 Gene Expression by Sex and ICU Status", 
       x = "Sex and ICU Status", 
       y = "Gene Expression", color= "Sex and ICU Status (sex.icu_status)") +
  theme_minimal()
```



